---
title: "Course Project 1"
output: html_document
---

```{r setting_global_options, echo=FALSE}
library(knitr)
opts_chunk$set(message = FALSE)
```

First, let's import the packages we are going to use. 

```{r}
library(data.table)
library(dplyr)
library(ggplot2)
library(hms)
library(lubridate)
library(stringr)
```

The **data.table** and **dplyr** packages are important to manipulate datasets 
more efficiently. The **ggplot2** package is to plot using the homonymous plotting
system. The **hms** and **lubridate** packages help when working with time and
date variables whereas the **stringr** package is useful to deal with strings.


## Step One

Let's load the dataset as a `data.table` (using the `fread()` function) and 
change the date variable to a "Date" type instead of "character". It's important
to first set your working directory to the folder where the `.csv` file is using
the `setwd()` function.

```{r}
foo <- fread("activity.csv")
foo <- foo %>% 
  mutate(date = ymd(date)) %>% 
  as.data.table()
```


## Step Two

First we aggregate the number of steps by date (excluding missing values). 
Then we plot the histogram of the total number of steps.

```{r, fig.align = 'center', warning = FALSE}
dt <- foo[steps != "NA", .(total = sum(steps)), by = .(date)]
ggplot(data = dt) + aes(x = total) + geom_histogram()
```


## Step Three

The next step is to take the mean and median of the total number of steps taken
each day. 

```{r}
mean_median <- dt %>% 
  summarize(mean = mean(total),
            median = median(total))
mean_median
```


## Step Four

Plotting a time series of the average number of steps taken in the 5-minute
interval:

```{r, fig.align = 'center', fig.width = 10}

foo <- foo %>% 
  mutate(interval = interval %>% as.character %>% str_pad(width = 4, pad = "0"))

foo <- foo %>% 
  mutate(interval = paste(str_sub(interval, 1, 2), str_sub(interval, 3, 4), sep = ":")) %>% 
  mutate(interval = parse_hm(interval)) %>% 
  mutate(interval = as.POSIXct(interval)) %>%
  as.data.table()

dt <- foo[steps != "NA", .(mean = mean(steps)), by = .(interval)]
ggplot(data = dt, aes(x = interval, y = mean)) + 
  geom_point() +
  geom_line() + 
  scale_x_datetime(breaks = scales::date_breaks("1 hour"),
                   date_labels = "%H:%M")
```


## Step Five

From the plot, it's possible to see that the 5-minute interval that, on average,
contains the maximum number of steps is after 08:30 and before 09:00.
To know the exact interval, we can do the following:

```{r}
dt <- dt %>% mutate(interval = as.hms(interval, tz = 'UTC')) %>% as.data.table()
dt[order(-mean)][1,]
```


## Step Six

### Describing

Checking how many rows exist with missing values in this dataset:
```{r}
missing_rows <- foo[!complete.cases(foo)] %>% nrow
```

Thus, from `r nrow(foo)` rows, there are `r missing_rows` with missing values. 
Which is about `r paste0(round(100*(missing_rows/nrow(foo))), "%")` of the 
dataset.

We can verify which variables are responsible for these missing values
```{r}
prop_NA <- sapply(foo, function(x) mean(is.na(x)))
prop_NA
```


Only the `steps` variable has missing values.
We can also investigate the proportion of missing values in each day.

```{r}
prop_NA_by_day <- foo[, .(prop_NA = mean(is.na(steps))), by = .(date)]
prop_NA_by_day
```

When the missing values appear, they do appear throught every interval of the day.  


### Imputing missing data

There are multiples ways to imput missing data. 
Here we will use the mean of the steps for the complete period to address the missing data issue.

```{r, warning = FALSE}
bar <- copy(foo)
bar[is.na(steps), steps := mean(foo$steps, na.rm = T)]
bar
```


## Step Seven 

Histogram of the total number of steps taken each day after missing values are imputed

```{r, fig.align = 'center'}
dt2 <- bar[, .(total = sum(steps)), by = .(date)]
ggplot(data = dt2) + aes(x = total) + geom_histogram()

```


## Step Eight

Let's investigate if there is any change on the activity pattern between 
weekdays and weekends before and after imputation.

We first need to create a variable that indicates if that date is a weekday or a
weekend.

```{r}
foo <- foo %>% 
  mutate(type_of_day = wday(date, label = T)) %>% 
  mutate(type_of_day = ifelse(type_of_day %in% c("Sat", "Sun"),
                              "Weekend", "Weekday")) %>% 
  as.data.table()
```


Now we plot the average number of steps taken per 5-minute interval comparing
weekdays and weekends before the imputation.

```{r, fig.align = 'center', fig.width = 10}
dt <- foo[steps != "NA", .(mean = mean(steps)), by = .(interval, type_of_day)]
ggplot(data = dt, aes(x = interval, y = mean)) + 
  geom_point() +
  geom_line() + 
  facet_grid(rows = vars(type_of_day)) + 
  scale_x_datetime(breaks = scales::date_breaks("1 hour"),
                   date_labels = "%H:%M")
```                 


We can also do the same thing to the dataset after imputation of missing values.

First we create the variable to distingish if the day is a weekday or a weekend.

```{r}
bar <- bar %>% 
  mutate(type_of_day = wday(date, label = T)) %>% 
  mutate(type_of_day = ifelse(type_of_day %in% c("Sat", "Sun"),
                              "Weekend", "Weekday")) %>% 
  as.data.table()
```


Now we plot the average number of steps taken per 5-minute interval comparing
weekdays and weekends after the imputation.

```{r, fig.align = 'center', fig.width = 10}
dt2 <- bar[, .(mean = mean(steps)), by = .(interval, type_of_day)]
ggplot(data = dt2, aes(x = interval, y = mean)) + 
  geom_point() +
  geom_line() + 
  facet_grid(rows = vars(type_of_day)) + 
  scale_x_datetime(breaks = scales::date_breaks("1 hour"),
                   date_labels = "%H:%M")
```                 


From the previous plots, we can say that during weekends people do not seem to 
take much steps from 06:00 to 08:00. Additionally, the average steps taken from 
08:00 to 21:00 has a smaller variance on weekends when compared to weekdays.

    